---
title: "Assignment 2 - Language Development in ASD - Part 1 - Explaining development"
author: "Lina Pedersen, Thea Pedersen, Sophie Korch, Andreas Jeppesen"
date: "16/9/2020"
output: html_document
---
    
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(include = FALSE)
```

# Assignment 2

In this assignment you will have to discuss a few important questions (given the data you have). More details below. The assignment submitted to the teachers consists of:
- a report answering and discussing the questions (so we can assess your conceptual understanding and ability to explain and critically reflect)
- a link to a git repository with all the code (so we can assess your code)

Part 1 - Basic description of language development
- Describe your sample (n, age, gender, clinical and cognitive features of the two groups) and critically assess whether the groups (ASD and TD) are balanced
- Describe linguistic development (in terms of MLU over time) in TD and ASD children (as a function of group). 
- Describe how parental use of language (in terms of MLU) changes over time. What do you think is going on?
- Include individual differences in your model of language development (in children). Identify the best model.

Part 2 - Model comparison
- Discuss the differences in performance of your model in training and testing data
- Which individual differences should be included in a model that maximizes your ability to explain/predict new data?
- Predict a new kid's performance (Bernie) and discuss it against expected performance of the two groups

Part 3 - Simulations to plan a new study
- Report and discuss a power analyses identifying how many new kids you would need to replicate the results

The following involves only Part 1.

## Learning objectives

- Summarize and report data and models
- Critically apply mixed effects (or multilevel) models
- Explore the issues involved in feature selection


# Quick recap
Autism Spectrum Disorder is often related to language impairment. However, this phenomenon has not been empirically traced in detail:
i) relying on actual naturalistic language production,  ii) over extended periods of time.

We therefore videotaped circa 30 kids with ASD and circa 30 comparison kids (matched by linguistic performance at visit 1) for ca. 30 minutes of naturalistic interactions with a parent. We repeated the data collection 6 times per kid, with 4 months between each visit. We transcribed the data and counted: 
i) the amount of words that each kid uses in each video. Same for the parent.
ii) the amount of unique words that each kid uses in each video. Same for the parent.
iii) the amount of morphemes per utterance (Mean Length of Utterance) displayed by each child in each video. Same for the parent. 

This data is in the file you prepared in the previous class. 

NB. A few children have been excluded from your datasets. We will be using them next week to evaluate how good your models are in assessing the linguistic development in new participants.

This RMarkdown file includes 
1) questions (see above). Questions have to be answered/discussed in a separate document that you have to directly submit on Blackboard.
2) A break down of the questions into a guided template full of hints for writing the code to solve the exercises. Fill in the code and the paragraphs as required. Then report your results in the doc for the teachers.

REMEMBER that you will have to have a github repository for the code and submit the answers to Blackboard without code (but a link to your github/gitlab repository). This way we can check your code, but you are also forced to figure out how to report your analyses :-)

Before we get going, here is a reminder of the issues you will have to discuss in your report:

1- Describe your sample (n, age, gender, clinical and cognitive features of the two groups) and critically assess whether the groups (ASD and TD) are balanced
2- Describe linguistic development (in terms of MLU over time) in TD and ASD children (as a function of group). 
3- Describe how parental use of language (in terms of MLU) changes over time. What do you think is going on?
4- Include individual differences in your model of language development (in children). Identify the best model.

# Let's go

### Loading the relevant libraries

Load necessary libraries : what will you need?
- e.g. something to deal with the data
- e.g. mixed effects models
- e.g. something to plot with

```{r Load Libraries, include = FALSE}
pacman::p_load(tidyverse, ggplot2, lme4, lmerTest, janitor)
```

### Define your working directory and load the data
If you created a project for this class and opened this Rmd file from within that project, your working directory is your project directory.

If you opened this Rmd file outside of a project, you will need some code to find the data:
- Create a new variable called locpath (localpath)
- Set it to be equal to your working directory
- Move to that directory (setwd(locpath))
- Load the data you saved last time (use read_csv(fileName))

```{r Load Data, include = FALSE}
# Loading data:
df <- read_csv("autismdata.csv")

```

### Characterize the participants (Exercise 1)

Identify relevant variables: participants demographic characteristics, diagnosis, ADOS, Verbal IQ, Non Verbal IQ, Socialization, Visit, Number of words used, Number of unique words used, mean length of utterance in both child and parents.

Make sure the variables are in the right format.

Describe the characteristics of the two groups of participants and whether the two groups are well matched.

```{r descriptive stats, include = FALSE}
pacman::p_load(purrr)

# Checking format of variables
str(df)

df %>% 
  mutate(
    SUBJ <- as.factor(SUBJ),           #Changing formats of some varaibles
    Diagnosis <- as.factor(Diagnosis),
    Ethnicity <- as.factor(Ethnicity),
    Gender <- as.factor(Gender)
    ) %>% 
  split(df$Diagnosis) %>%    # splitting the dataframe into two based on Diagnosis
  map(summary)            # Mapping the function summary() onto each dataframe to evaluate how balanced the two groups are




```

The sample included mostly young (<20) white males ...

[REPORT THE RESULTS]

## Let's test hypothesis 1: Children with ASD display a language impairment  (Exercise 2)

### Hypothesis: The child's MLU changes: i) over time, ii) according to diagnosis

Let's start with a simple mixed effects linear model

Remember to plot the data first and then to run a statistical test.
- Which variable(s) should be included as fixed factors?
- Which variable(s) should be included as random factors?

```{r ex2, include = FALSE}
# Plotting data:
df %>% 
  ggplot() +
  aes(x=VISIT, y=CHI_MLU, color = Diagnosis) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_classic()


# Which random/fixed effects to include?
# Fixed effects should be Age and Diagnosis, because they are the main effect we are interested in
# Random effect should be SUBJ because we are not interested in the individual differences
# Dependent variable is the children MLU

# We are going with a linear mixed effects model including both IVs and by-Subject random slopes for Visit nr and random intercept. Model would not converge propperly if we also allow by-subject random slope for Diagnosis.
msimple <- lmerTest::lmer(CHI_MLU ~ VISIT * Diagnosis + (1 + VISIT |SUBJ) , data = df , REML = FALSE)

summary(msimple)

```

How would you evaluate whether the model is a good model?

```{r ex2 evaluate, include = FALSE}
# Checking the marginal and conditional R-squared of the model:

pacman::p_load(MuMIn)

# Getting marginal and conditional R-squared
r.squaredGLMM(msimple)

```

Not too good, right? Let's check whether a growth curve model is better.
Remember: a growth curve model assesses whether changes in time can be described by linear, or quadratic, or cubic (or... etc.) components.
First build the different models, then compare them to see which one is better.

```{r ex2 growth curve, include = FALSE}
# We make a new model with interction between Visit and Diagnosis AND a growth curve for Visit
full <- lmerTest::lmer(CHI_MLU ~ VISIT * Diagnosis + I(VISIT^2) + (1 + VISIT|SUBJ) , data = df, REML = FALSE)

summary(full)

# Comparing the new full model to the simple model from before:
AIC(msimple , full)

# The full model is slightly better than the simple model (AIC = 545.9 vs. AIC = 572.5)
```

Exciting right? Let's check whether the model is doing an alright job at fitting the data. Plot the actual CHI_MLU data against the predictions of the model fitted(model). 

```{r}
fitted(full) # This function gets the predicted values from the full-model

predicted.df <- data.frame(fitted(full)) # saving the values as a df

CHI_MLU <- df$CHI_MLU %>%  # saving the real CHI_MLU vlaues
  na.exclude()             # in which we exclude the NA's


predicted.df$CHI_MLU <- CHI_MLU   # Adding a column to the predicted.df with the CHI_MLU values

# Making a scatterplot in which we compare the actual values of CHI_MLU (x-axis) to the values of CHI_MLU predicted by the model (y-axis)
predicted.df %>% 
ggplot() + 
  aes(x = CHI_MLU , y = fitted.full. ) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "Red") +
  theme_classic()

# The points fall (roughly) on a linear diagonal, indicating that the model is good at predicting the true values of CHI_MLU.
# We can see that there is no bias as the points do not tend to fall equally on either side of the ideal diagonal


```

Now it's time to report our results.
Remember to report:
- the estimates for each predictor (beta estimate, standard error, p-value)
- A plain word description of the results
- A plot of your model's predictions (and some comments on whether the predictions are sensible)

[REPORT THE RESULTS]
Linguistic development of children MLU is affected by ... [COMPLETE]

## Let's test hypothesis 2: Parents speak equally to children with ASD and TD  (Exercise 3)

### Hypothesis: Parental MLU changes: i) over time, ii) according to diagnosis

```{r ex3, include = FALSE}
# Plotting data:
df %>% 
  ggplot() +
  aes(x=VISIT, y=MOT_MLU, color = Diagnosis) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_classic()


#fitting a simple model like the one from the Child's MLU
msimple2 <- lmerTest::lmer(MOT_MLU ~ VISIT * Diagnosis + (1 + VISIT |SUBJ) , data = df , REML = FALSE)

summary(msimple2)

#fitting a full model like the one from the child's MLU
full2 <- lmerTest::lmer(MOT_MLU ~ VISIT * Diagnosis + I(VISIT^2) + (1 + VISIT|SUBJ) , data = df, REML = FALSE)

summary(full2)

```

Parent MLU is affected by ... but probably not ...
[REPORT THE RESULTS]

### Adding new variables (Exercise 4)

Your task now is to figure out how to best describe the children linguistic trajectory. The dataset contains a bunch of additional demographic, cognitive and clinical variables (e.g.verbal and non-verbal IQ). Try them out and identify the statistical models that best describes your data (that is, the children's MLU). Describe how you selected the best model and send the code to run the model to Victor and Byurakn.


```{r ex4, include = FALSE}
# We already included individual differences in our model, but here is a model wihtout random effect for Subject for comparison
m3 <- lm(CHI_MLU ~ VISIT * Diagnosis + I(VISIT^2), data = df)

# We compare the AIC of the two:
AIC(full, m3)

# The full model for explaining Child MLU is superior to a model without random effect for subject (AIC = 545.9 vs. AIC = 794.2)

# We build additional models with the other parameters:

# Model including effect of Social Skills
m4 <- lmerTest::lmer(CHI_MLU ~ VISIT * Diagnosis + I(VISIT^2) + (1 + VISIT|SUBJ) + SocialSkills1, data = df, REML = FALSE) # failed to converge
isSingular(m4)  # It is not singular

# Model with Verbal IQ
m5 <- lmerTest::lmer(CHI_MLU ~ VISIT * Diagnosis + I(VISIT^2) + (1 + VISIT|SUBJ) + VerbalIQ1, data = df, REML = FALSE)

# Model with non-verbal IQ
m6 <- lmerTest::lmer(CHI_MLU ~ VISIT * Diagnosis + I(VISIT^2) + (1 + VISIT|SUBJ) + NonVerbalIQ1, data = df, REML = FALSE)

# Model with Gender
m7 <- lmerTest::lmer(CHI_MLU ~ VISIT * Diagnosis + I(VISIT^2) + (1 + VISIT|SUBJ) + Gender , data = df, REML = FALSE) # failed to converge
isSingular(m7) # It is not singular

# Model with Severity of symptoms
m8 <- lmerTest::lmer(CHI_MLU ~ VISIT * Diagnosis + I(VISIT^2) + (1 + VISIT|SUBJ) + SeverityOfSymptoms1, data = df, REML = FALSE)


# Comparing all the models with AIC
AIC(full, m4, m5, m6, m7, m8)

# The best model is the full model with Verbal IQ as an additional parameter

summary(m5)
# Summarizing the model reveals that the effect of verbal IQ is significant

```

In addition to ..., the MLU of the children is also correlated with ...
Using AIC / nested F-tests as a criterium, we compared models of increasing complexity and found that ...

[REPORT THE RESULTS]